---
title: "Eigenvector-Based Semiparametric Filtering of Spatial Autocorrelation in Regression Models"
date: "March 17, 2021"
output:
  beamer_presentation:
    includes:
      in_header:
        - pres_template/pres_template.sty
        - pres_template/pkg.tex
        - title.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Motivation

* spatial misspecification can lead to inefficient or even biased and inconsistent parameter estimates in regression models
* social scientists primarily deal with a spatial misspecification problem in one of two ways:
  1. ignore it altogether and assume \textit{iid} observations
  2. apply (parametric) spatial regression models

\vspace{.3cm}

* semiparametric filtering techniques can offer an attractive alternative
  * ease of estimation (standard OLS or ML estimators)
  * straightforward interpretation
  * accounts for spatial autocorrelation at different scales/resolutions
  * easily adaptable to GLMs


## Spatial Misspecification in Linear Models

\vspace{-.4cm}

\[
\bm{y} = \bm{X\beta} + \bm{e}
\]

\pause
\vspace{-.2cm}

\begin{minipage}[t]{.3\textwidth}
\centering
\underline{Spatial Error DGP}
\footnotesize
\[
\bm{y} = \bm{X\beta} + \underbrace{(\bm{I}-\rho\bm{W})^{-1}\bm{\epsilon}}_{\bm{e}_{SEM}}
\]
\end{minipage}
\begin{minipage}[t]{.28\textwidth}
\centering
\underline{Spatial-X DGP}
\footnotesize
\[
\bm{y} = \bm{X\beta} + \underbrace{\rho\bm{WX}+\bm{\epsilon}}_{\bm{e}_{SLX}}
\]
\end{minipage}
\begin{minipage}[t]{.4\textwidth}
\centering
\underline{Spatial Lag DGP}
\footnotesize
\begin{align}
\nonumber\bm{y} & = \bm{X\beta} + \underbrace{\rho\bm{Wy}+\bm{\epsilon}}_{\bm{e}_{SAR}}\\
\nonumber\bm{y} & = (\bm{I}-\rho\bm{W})^{-1} (\bm{X\beta} + \bm{\epsilon})
\end{align}
\end{minipage}

\pause
<!-- \vspace{-.5cm} -->
\footnotesize
\[
(\bm{I}-\rho\bm{W})^{-1} = (\bm{I} + \rho\bm{W}+\rho^2\bm{W}+...)
\]


\vspace{-.7cm}

\begin{minipage}[t]{.3\textwidth}
\footnotesize
\[\begin{array}{rl}
\bm{y} = & \bm{X\beta}\\
&\left.\begin{array}{l} + \rho\bm{W}\bm{\epsilon}\\
+ \rho^2\bm{W}^2\bm{\epsilon}\\
+ ... + \bm{\epsilon}\end{array}\right\}\bm{\epsilon}_{SEM}
\end{array}
\]
\end{minipage}
\begin{minipage}[t]{.28\textwidth}
\footnotesize
\
\end{minipage}
\begin{minipage}[t]{.4\textwidth}
\footnotesize
\[\begin{array}{rl}
\bm{y} = & \bm{X\beta}\\
&\left.\begin{array}{l} + \rho\bm{W}(\bm{X\beta}+\bm{\epsilon})\\
+ \rho^2\bm{W}^2(\bm{X\beta}+\bm{\epsilon})\\
+ ... + \bm{\epsilon}\end{array}\right\}\bm{\epsilon}_{SAR}
\end{array}
\]
\end{minipage}

\pause
<!-- \vspace{.2cm} -->

\begin{minipage}[t]{.325\textwidth}
\centering
\footnotesize
\underline{$\bm{e}_{SEM}$}
\vspace{-.3cm}
\[
\underbrace{\sum_{r=1}^{\infty}\rho^r\bm{W}^r\bm{\epsilon}}_{\text{spatial part}} + \underbrace{\vphantom{\sum_{i}^{i}}\bm{\epsilon}}_{\text{noise}}
\]
\end{minipage}
\begin{minipage}[t]{.325\textwidth}
\centering
\footnotesize
\underline{$\bm{e}_{SLX}$}
\vspace{-.3cm}
\[
\underbrace{\vphantom{\sum_{r=1}^{\infty}}\rho\bm{WX}}_{\text{spatial part}} +\underbrace{\vphantom{\sum_{r=1}^{\infty}}\bm{\epsilon}}_{\text{noise}}
\]
\end{minipage}
\begin{minipage}[t]{.325\textwidth}
\centering
\footnotesize
\underline{$\bm{e}_{SAR}$}
\vspace{-.3cm}
\[
\underbrace{\sum_{r=1}^{\infty}\rho^r\bm{W}^r(\bm{X\beta} + \bm{\epsilon})}_{\text{spatial part}} + \underbrace{\vphantom{\sum_{i}^{i}}\bm{\epsilon}}_{\text{noise}}
\]
\end{minipage}


## Eigenvector-Based Spatial Filtering

\underline{Intuition:}

ESF partitions the response variable into i) a systematic trend, ii) a stochastic spatial signal, and iii) \textit{iid} disturbances

\vspace{.7cm}

\underline{Example: SAR DGP}

\begin{center}
\begin{tikzpicture}[sibling distance=10em]
\node{\begin{tabular}{c}$\bm{y}$\\{\tiny(outcome)}\end{tabular}}
	child[->]{node{\begin{tabular}{c}$\bm{X\beta}$\\{\tiny(systematic trend)}\end{tabular}}}
	child[->]{node{\begin{tabular}{c}$\sum_{r=1}^{\infty}\rho^r\bm{W}^r(\bm{X\beta} + \bm{\epsilon})+\epsilon$\\{\tiny(autocorrelated part)}\end{tabular}}
		child[->]{node{\begin{tabular}{c}$\sum_{r=1}^{\infty}\rho^r\bm{W}^r(\bm{X\beta} + \bm{\epsilon})$\\{\tiny(stochastic spatial signal)}\end{tabular}}}
		child[->]{node{\begin{tabular}{c}$\epsilon$\\{\tiny(\textit{iid} disturbances)}\end{tabular}}}};
\end{tikzpicture}
\end{center}


## Spatial Eigenfunction Analysis

ESF is based on the spectral decomposition of a centered (and symmetric/ symmetrized) connectivity matrix: $\bm{W}$:
\[
\bm{MWM} = \bm{E\Lambda E}^{-1} = \bm{E\Lambda E}'
\]

* demeaning projector $\bm{M}=(\bm{I}-\bm{11}'/n)$ -- also ensures that all eigenvectors are orthogonal and uncorrelated
* $\bm{E}$ are all $n$ eigenvectors
* $\bm{\Lambda}$ is a diagonal matrix of the corresponding eigenvalues $\bm{\lambda}$


\textcolor{white}{Direct relationship to the numerator of the global Moran coefficient:}

\[
\textcolor{white}{MC(\bm{x}) = \frac{n}{\bm{1}'\bm{W1}} \frac{\bm{x}'\bm{MWM}x}{\bm{x}'\bm{Mx}}}
\]

## Spatial Eigenfunction Analysis

ESF is based on the spectral decomposition of a centered (and symmetric/ symmetrized) connectivity matrix: $\bm{W}$:
\[
\bm{\textcolor{red}{MWM}} = \bm{E\Lambda E}^{-1} = \bm{E\Lambda E}'
\]

* demeaning projector $\bm{M}=(\bm{I}-\bm{11}'/n)$ -- also ensures that all eigenvectors are orthogonal and uncorrelated
* $\bm{E}$ are all $n$ eigenvectors
* $\bm{\Lambda}$ is a diagonal matrix of the corresponding eigenvalues $\bm{\lambda}$


Direct relationship to the numerator of the global Moran coefficient:

\[
MC(\bm{x}) = \frac{n}{\bm{1}'\bm{W1}} \frac{\bm{x}'\bm{\textcolor{red}{MWM}x}}{\bm{x}'\bm{Mx}}
\]


## Eigenvectors as Synthetic Map Patterns

* eigenvectors $\bm{E}$ depict distinct -- and uncorrelated -- synthetic map patterns
* corresponding eigenvalues indicate the level of SA

```{r echo=F, include=F}
rm(list=ls())
library(spfilteR)
data("fakedata")

# eigenfunction decomposition
EVs <- getEVs(W=W,covars=NULL)
E <- EVs$vectors

# preliminaries for plot
library(RColorBrewer)
library(lattice)
col <- colorRampPalette(brewer.pal(n=9, "BuPu"))(400)
coord.x <- coord.y <- seq_len(sqrt(nrow(W)))
coord <- expand.grid(x=coord.x,y=coord.y)
```

\vspace{-.3cm}
\pause

\begin{minipage}[t]{.32\textwidth}
\centering
\tiny
Eigenvector 1\\
(MC = `r round(EVs$moran[1],3)`)
```{r out.width="80%",out.height="80%"}
lattice::levelplot(E[,1]~coord[,1]*coord[,2],col.regions=col,colorkey=F
                   ,contour=F,xlab=NULL,ylab=NULL,scales=list(draw=F))
```
\end{minipage}
\pause
\begin{minipage}[t]{.32\textwidth}
\centering
\tiny
Eigenvector 10\\
(MC = `r round(EVs$moran[10],3)`)
```{r out.width="80%",out.height="80%"}
lattice::levelplot(E[,10]~coord[,1]*coord[,2],col.regions=col,colorkey=F
                   ,contour=F,xlab=NULL,ylab=NULL,scales=list(draw=F))
```
\end{minipage}
\pause
\begin{minipage}[t]{.32\textwidth}
\centering
\tiny
Eigenvector 20\\
(MC = `r round(EVs$moran[20],3)`)
```{r out.width="80%",out.height="80%"}
lattice::levelplot(E[,20]~coord[,1]*coord[,2],col.regions=col,colorkey=F
                   ,contour=F,xlab=NULL,ylab=NULL,scales=list(draw=F))
```
\end{minipage}

\begin{minipage}[t]{.32\textwidth}
\centering
\tiny
Eigenvector 44\\
(MC = `r round(EVs$moran[44],3)`)
```{r out.width="80%",out.height="80%"}
lattice::levelplot(E[,44]~coord[,1]*coord[,2],col.regions=col,colorkey=F
                   ,contour=F,xlab=NULL,ylab=NULL,scales=list(draw=F))
```
\end{minipage}
\begin{minipage}[t]{.32\textwidth}
\centering
\tiny
Eigenvector 80\\
(MC = `r round(EVs$moran[80],3)`)
```{r out.width="80%",out.height="80%"}
lattice::levelplot(E[,80]~coord[,1]*coord[,2],col.regions=col,colorkey=F
                   ,contour=F,xlab=NULL,ylab=NULL,scales=list(draw=F))
```
\end{minipage}
\pause
\begin{minipage}[t]{.32\textwidth}
\centering
\tiny
MCs of all\\
Eigenvectors
```{r out.width="80%",out.height="80%"}
par(oma=c(4,4,0,0),mgp = c(3, 2, 0))
plot(0,ylim=c(min(EVs$moran),max(EVs$moran)),xlim=c(1,length(EVs$moran))
     ,axes=F,ann=F,type="n",las=1)
  # EVs
  points(y=EVs$moran,x=1:length(EVs$moran),pch=16,cex=1)
  abline(h=0,lty=2,cex=.5)
  # axes & labels
  axis(1,cex.axis=3)
  axis(2,cex.axis=3)
  mtext("Eigenvectors",side=1,line=.5,outer=T,cex=5)
  mtext("MC",side=2,line=.5,outer=T,cex=5)
```
\end{minipage}


## Real-World Map Patterns

* $\bm{E}$ depict \underline{all} possible spatial patterns permitted by $\bm{W}$
* more complex patterns can be obtained by a linear combination of eigenvectors

```{r echo=F, include=F}
rm(list=ls())
library(spdep)

# define color palette
col <- brewer.pal(n=9, "Blues")

# get German NUTS 3 regions shapefile & data
load("../SEA/Data/eurostat2017.RData")
ger3 <- st_read(dsn="../SEA/Shapefile/Germany")

# specify spatial matrix
nb <- poly2nb(ger3$geometry, queen=TRUE, row.names=ger3$NUTS_ID)
sty <- "W"
W <- nb2mat(nb,style=sty)

# ESF
var <- data$medianage
MCobs <- MI.vec(x=var,W=W)
filter <- lmFilter(y=var,W=W,objfn="MI")
MCfil <- filter$moran
```

\begin{minipage}[t]{.32\textwidth}
\centering
\tiny
Median Age\\
```{r out.width="80%",out.height="80%"}
# colcode
seq <- quantile(var,probs=c(0,1/9,2/9,3/9,4/9,5/9,6/9,7/9,8/9))
colcode <- sapply(var,function(x) sum(x >=seq))
# map
par(oma=c(0,0,0,0),mar=c(0,0,0,0))
plot(st_geometry(ger3),col=col[colcode])
```

MC = `r round(MCobs$I,3)`\
(\textit{p} = `r round(MCobs$pI,3)`)
\end{minipage}
\pause
\begin{minipage}[t]{.32\textwidth}
\centering
\tiny
Spatial Filter\\
```{r out.width="80%",out.height="80%"}
# colcode
seq <- quantile(filter$other$sf,probs=c(0,1/9,2/9,3/9,4/9,5/9,6/9,7/9,8/9))
colcode <- sapply(filter$other$sf,function(x) sum(x >=seq))
# map
par(oma=c(0,0,0,0),mar=c(0,0,0,0))
plot(st_geometry(ger3),col=col[colcode])
```

MC = `r round(MCfil["Initial","Observed"],3)`\
(\textit{p} = `r round(MCfil["Initial","p-value"],3)`)
\end{minipage}
\begin{minipage}[t]{.32\textwidth}
\centering
\tiny
Filtered Residuals\\
```{r out.width="80%",out.height="80%"}
# colcode
seq <- quantile(filter$residuals[,"Filtered"],probs=c(0,1/9,2/9,3/9,4/9,5/9,6/9,7/9,8/9))
colcode <- sapply(filter$residuals[,"Filtered"],function(x) sum(x >=seq))
# map
par(oma=c(0,0,0,0),mar=c(0,0,0,0))
plot(st_geometry(ger3),col=col[colcode])
```

MC = `r round(MCfil["Filtered","Observed"],3)`\
(\textit{p} = `r round(MCfil["Filtered","p-value"],3)`)
\end{minipage}


## Eigenvector Selection for the Spatial Filter

However, it is impossible to include all $n$ eigenvectors as regressors

1. Identification of a candidate set $\bm{E}^C \subset \bm{E}$ based on
    - sign of SA
    - strength of SA
2. select relevant map patterns $\bm{E}^* \subset \bm{E}^C$ using supervised or unsupervised stepwise regression

\vspace{-.35cm}

\begin{align}
\nonumber \bm{y} &= \bm{X\beta} + \overbrace{\bm{E}\bm{\gamma} + \vphantom{\gamma}\bm{\epsilon}}^{\bm{e}_{OLS}}\\
\nonumber &\approx \bm{X\beta} + \underbrace{\bm{E}^*\bm{\gamma}}_{filter} + \underbrace{\vphantom{\gamma}\bm{\epsilon}}_{noise}
\end{align}

* different criteria can be used to select eigenvectors, e.g.:
  * maximization of model fit (e.g., AIC, BIC, $R^2$)
  * significance level of eigenvectors
  * minimization of residual SA

## \ 

\centering
(Preliminary)

\LARGE\bf Monte Carlo Evidence


## Reduction in Residual SA

```{r echo=F, include=F}
rm(list=ls())
library(RColorBrewer)

load("../Results.RData")

sar <- matrix(NA,nrow=nrow(input),ncol=ncol(sim_out[[1]]$sar))
colnames(sar) <- names(sim_out[[1]]$sar)
for(i in 1:nrow(input)){
  sar[i,] <- as.matrix(sim_out[[i]]$sar)
}
```

```{r fig.width=5.5, fig.height=3.5}
brewcol <- brewer.pal(n=4,name="RdGy") # "PuOr" - "RdGy" - "RdBu"
layout(matrix(c(1,2,3),ncol=3,byrow=F))
par(oma=c(3,3.5,2,3.5),mar=c(.3,.3,.3,.3))
for(w in unique(input$W_id)){
  sel <- input$W_id==w & input$sd.e==1
  plot(0,xlim=c(0,max(input$p)),ylim=c(-1,30),type="n",axes=F,ann=F)
  # initial MI
  lines(x=unique(input$p),y=c(median(sar[sel&input$p==0,"moran_init"])
                              ,median(sar[sel&input$p==.25,"moran_init"])
                              ,median(sar[sel&input$p==.5,"moran_init"])
                              ,median(sar[sel&input$p==.75,"moran_init"]))
        ,type="b",pch=21,col=brewcol[1],bg=brewcol[1])
  # Filtered (p)
  lines(x=unique(input$p),y=c(median(sar[sel&input$p==0,"moran_filtered_p"])
                              ,median(sar[sel&input$p==.25,"moran_filtered_p"])
                              ,median(sar[sel&input$p==.5,"moran_filtered_p"])
                              ,median(sar[sel&input$p==.75,"moran_filtered_p"]))
        ,lty=4,type="b",pch=22,col=brewcol[2],bg=brewcol[2])
  # Filtered (MC)
  lines(x=unique(input$p),y=c(median(sar[sel&input$p==0,"moran_filtered_MI"])
                              ,median(sar[sel&input$p==.25,"moran_filtered_MI"])
                              ,median(sar[sel&input$p==.5,"moran_filtered_MI"])
                              ,median(sar[sel&input$p==.75,"moran_filtered_MI"]))
        ,lty=3,type="b",pch=23,col=brewcol[3],bg=brewcol[3])
  # Filtered (R2)
  lines(x=unique(input$p),y=c(median(sar[sel&input$p==0,"moran_filtered_R2"])
                              ,median(sar[sel&input$p==.25,"moran_filtered_R2"])
                              ,median(sar[sel&input$p==.5,"moran_filtered_R2"])
                              ,median(sar[sel&input$p==.75,"moran_filtered_R2"]))
        ,lty=2,type="b",pch=24,col=brewcol[4],bg=brewcol[4])
  if(w==min(input$W_id)){
    axis(2,at=c(0,5,10,15,20,25,30,35),las=1,mgp=c(3,.7,0))
    legend("topleft",inset=0
           ,legend=c(expression("OLS"[" "]),expression("ESF"[rho]),expression("ESF"[MC]),expression("ESF"[R^2]))
           ,col=brewcol
           ,pch=21:24
           ,text.width=strwidth(expression("ESF"[MC]))*.9
           ,pt.bg=brewcol
           ,cex=.7
           ,y.intersp=1.1)
  }
  
  axis(1,at=unique(input$p),labels=c("0","0.25","0.5","0.75"),mgp=c(3,.5,0))
  
  if(w==max(input$W_id)) axis(4,at=c(0,5,10,15,20,25,30,35),las=1,mgp=c(3,.7,0))
  if(w==1) mtext(expression(bold(W)["50"]),side=3,line=.5)
  if(w==2) mtext(expression(bold(W)["200"]),side=3,line=.5)
  if(w==3) mtext(expression(bold(W)["435"]),side=3,line=.5)
  box("plot")
}
mtext(expression(italic(rho)),side=1,line=2,outer=T)
mtext("Standardized Moran Coefficient",side=2,line=2,outer=T)
```


## Improved Model Fit

```{r echo=F, include=F}
rm(list=ls())
load("../Res.RData")

sar <- matrix(NA,nrow=nrow(input),ncol=ncol(sim_out[[1]]$sar))
colnames(sar) <- names(sim_out[[1]]$sar)
for(i in 1:nrow(input)){
  sar[i,] <- as.matrix(sim_out[[i]]$sar)
}
```

```{r fig.width=5.5, fig.height=3.5}
sel <- input$sd.e==1 & input$W_id==3
layout(matrix(c(1,2,3,4),ncol=2,byrow=T))
par(oma=c(0,3,0,0),mar=c(1.5,1,1,2))
for(p in unique(input$p)){
  plot(0,xlim=c(1,4),ylim=c(.2,1),type="n",axes=F,ann=F)
  text(x=2.5,y=.9,labels=bquote(rho~"="~.(eval(p))),cex=1)
  for(i in 1:nsim){
    lines(x=c(1:4),y=c(sar[sel&input$p==p,"fit_init"][i]
                       ,sar[sel&input$p==p,"fit_filtered_p"][i]
                       ,sar[sel&input$p==p,"fit_filtered_MI"][i]
                       ,sar[sel&input$p==p,"fit_filtered_R2"][i])
          ,lwd=.02,col=rgb(0,0,0,alpha=.03))
  }
  m <- sort(sar[sel&input$p==p,"fit_init"],decreasing=F)[nsim*.5]
  lines(x=c(1:4),y=c(sar[sel&input$p==p,"fit_init"][sar[sel&input$p==p,"fit_init"]==m]
                     ,sar[sel&input$p==p,"fit_filtered_p"][sar[sel&input$p==p,"fit_init"]==m]
                     ,sar[sel&input$p==p,"fit_filtered_MI"][sar[sel&input$p==p,"fit_init"]==m]
                     ,sar[sel&input$p==p,"fit_filtered_R2"][sar[sel&input$p==p,"fit_init"]==m])
        ,lwd=2,col="navyblue") # or white?
  axis(1,at=c(1,2,3,4),labels=c("OLS",expression("ESF"[rho]),expression("ESF"[MC]),expression("ESF"[R^2]))
       ,col=NA,line=-1,cex.axis=.9) # ,col.ticks=1
  abline(v=c(1,2,3,4),col=rgb(0,0,0,alpha=.3),lwd=1)
  axis(2,at=c(.2,.4,.6,.8,1),las=1,cex.axis=.9,mgp=c(3,.7,0))
  mtext(expression("Adjusted R"^"2"),side=2,line=1.5,outer=T)
}
```


## Conclusion

* ESF offers numerous advantages for political scientists
* however, applicability depends on RQ (as always!)
* things to consider:

\begin{minipage}[t]{.495\textwidth}
\begin{center}
\underline{Pros}
\end{center}
\begin{itemize}
\item[+] ease of model estimation
\item[+] straightforward interpretation of parameters
\item[+] flexibility
\item[] {\footnotesize (no need to specify the spatial pattern in each variable)}
\item[+] generalizability
\item[] \footnotesize {(also applicable to GLMs -- with some modifications)}
\end{itemize}
\end{minipage}
\begin{minipage}[t]{.495\textwidth}
\begin{center}
\underline{Cons}
\end{center}
\begin{itemize}
\item[-] ``removes'' indirect spillovers
\item[-] computationally demanding for large N
\item[-] over- or undercorrection of SA possible
\end{itemize}
\end{minipage}


## Additional Resources

* \texttt{spfilteR} package:
\begin{itemize}
\item[] \underline{CRAN:} \url{https://CRAN.R-project.org/package=spfilteR}
\item[] \underline{GitHub:} \url{https://github.com/sjuhl/spfilteR}
\end{itemize}

* further projects \& working papers
  1. introductory paper on the \texttt{spfilteR} package (under review)
  2. project on Moran eigenvector maps and spatial eigenfunction analysis

\vspace{1cm}

\centering
**Feedback \& suggestions are highly appreciated!**\
\texttt{sebastian.juhl@gess.uni-mannheim.de}\
\url{www.sebastianjuhl.com}
